Data-Driven Modelling \& \\ Forecasting of Chaotic Systems

### 2022-01-17 draft_16Jan/forcomparison.tex
Grammarly changes copied in

Experimenting with biological, physical, and artificial systems to generate a more informative dynamical model is a well-established practice in modern science. Traditional methods for modelling physical systems are based on laws of physics that are based either on empirical relationships or intuition. For systems that evolve with time, physical laws yield mathematical equations that govern how the quantities evolve with time. However, our world is much more complex than that which can be distilled into elegant equations. We do not fully understand many of the more complex systems, nor do they provide us with a good physical intuition of the underlying principles governing system dynamics. To complicate this, the underlying systems we observe often display sensitive dependence on initial conditions despite having highly similar initial conditions. This is because differing orbits diverge quickly and to such an extent that it becomes seemingly impossible to retrace their steps back to the original conditions. Even the presence of usually' negligible' computational noise or measurement error renders long-term pointwise prediction infeasible. There are many difficulties that one encounters while modelling such systems:

\vspace{-8mm}
\begin{enumerate}[noitemsep, label=\roman*.]
  \item One may not have access to the complete states of the systems
  \item The system may be described by functions that behave wildly because their graphs have a wild oscillatory behaviour, i.e., a large functional complexity.
\end{enumerate}

@@

Examples in the literature attempting to forecast data from such systems have tried many different approaches with differing degrees of success. When we have complete access to states of the system, an ordinary differential equation can be obtained from data ([?, ?] and [?, ?]) wherein one could approximate the vector field by a library of functions to get interpretable models.
Examples in the literature attempting to forecast data from such systems have tried a number of different approaches with differing degrees of success. 
When we have complete access to states of the system, an ordinary differential equation can be obtained from data (\cite{brunton2016discovering, champion2019data} and \cite{small2002modeling,xu2006modeling}) wherein one could approximate the vector field by a library of functions to obtain interpretable models.

@@

In this chapter, we provide a brief description of what a discrete-time dynamical system is and what it means for it to exhibit chaos. We refer to \cite{devaney2018introduction, de2013elements} for more details.

@@

Formally, a function $T: U \to U$, where $U$ is some set is  a \emph{discrete-time dynamical system} and its iterates $\{u,Tu,T^2u,\ldots\}$, where $T^n$ denotes the $n$-fold composition of $T$ with itself, describe the evolution of an initial condition $u\in U$ (Note that we frequently drop the brackets and denote $T(u)$ by $Tu$ so as to simplify notations).

@@

Of course, discrete-time dynamical systems need not always arise through a differential equation, and once again we may consider the field of ecology where discrete dynamical systems are often directly derived or assumed. \textbf{Citation}
There is a school of thought that advocates discrete-time dynamical systems to be more natural for modelling real-world observations than differential equations and we refer the interested reader to \cite{saber2010introduction}.

@@

Two examples of invariant sets include a fixed point where $Tu=u$ for  $u\in U$, and a periodic orbit, i.e., a set of iterates $\{u,Tu, T^2u,\ldots,T^pu\}$, where $T^{p+1}u=u$ for some $p\in\mathbb{Z}$.  The entire space $U$ could be also be invariant.
Consider for example the space $U=[0,1]$, where $Tu=4u(1-u)$ and then $U$ is invariant (as every $u\in{U}$ can be written as $u = 4x(1-x)$ for some $x\in{U}$).

@@

However,  when the space $U$ is compact, all orbits approach an invariant set. This is since the set of limit points of the orbit can be shown to be invariant \cite{de2013elements}. So, when $U$ is compact,  the $\omega$-limit set $\omega(u;T)$ of a point $u$ defined to the collection of limit points of the sequence $\{x,Tu,T^2u,\ldots\}$ is nonempty, and $\omega(u;T)$ is invariant.

@@

Invariant sets have various properties. Invariant sets can be attracting or repelling depending on how orbits in its vicinity behave. 
Recall the examples $Tu =u/2$ and $Tu=2u$ defined on $\mathbb{R}$ (where $\{0\}$ ``attracts" orbits) or the example, $Tu=u^2$ defined on $[0,1]$ (where $\{1\}$ ``repels" orbits). We are interested in attractive invariant sets since they capture the long-term dynamics as time increases. 
In particular we are interested in those invariant sets named attractors.

@@

Indeed in our previous examples, it can be verified that the system  $Tu=u/2$ defined on  $\mathbb{R}$ and $Tu=u^2$ defined on $[0,1]$  the singleton set $\{0\}$ is an attractor.  For the system,  $Tu=1-|2u-1|$ on $[0,1]$, one may easily verify that the only attractor is the entire space $[0,1]$. This is since between any two points $u< v$ in $[0,1]$, and for any $a,b$ so that $u\le a < b \le v$, we can find an $n$ so that $T^n(a,b)=[0,1]$ (as would be explained later in this chapter).

A dynamical system can have several attractors, and may also be contained in another attractor. For the example, $Tu =u^2$ on $[0,1]$, both $\{0\}$ and $[0,1]$ are attractors. It is a known result that for dynamical systems on a compact space there always exists an attractor.

The dynamics restricted to an invariant set can be complicated. For instance an invariant set could be just a single point or it could have an infinite set.    If the invariant set is infinite then there is a possibility of complicated dynamics, and a particular well-studied phenomenon of such complexity gives rise to what is called  chaotic behaviour, and has been studied in detail over the last fifty years.

@@

In the 1960s, a number of mathematicians and mathematically interested scientists independently discovered chaos in the mathematical sense. The meteorologist Edward Lorenz may have been the first to explain this phenomenon in his 1963 paper \cite{lorenz1963deterministic}. The notions of invariance, attractivity and chaos may also be described for continuous systems, and Lorenz's system comprised of a system of differential equations. The narrative of Lorenz's discovery of chaos, as well as the history of other forerunners in this subject, are fascinating. For those who are interested, we highly recommend James Gleick's book Chaos: The Making of a New Science \cite{gleick2008chaos}. It clearly illustrates these experiences, explains why chaos was such a startling and crucial mathematical and scientific discovery, and describes the underlying mathematical notions for non-specialists.

There are several intricately different definitions of chaos available in the literature \textbf{(Cite)}, with each of them indicating some aspect of complexity.
In practice, it is not possible to verify which definition or notion does the underlying dynamical system satisfies when only data is observed from it. We do, however, specifically recall the definition of chaos in the sense of Devaney \cite{devaney2018introduction,de2013elements} so as to understand some nuances behind the complexity. Devaney's definition of chaos has three requirements, with the first being the notion  of sensitive dependence on initial conditions.

@@

It is important to acquire a sense of this definition as it can easily be misunderstood. It is common misconception to interpret SDIC as two close points ($u$ and $v$) that eventually become separated by a distance $\delta$ under iteration by $T$. But this is not true. To fully comprehend all the subtleties of the concept, one needs to discuss it more thoroughly by considering each sentence with care and attention, whereafter one may examine how they fit together to convey the concept of sensitive dependence.

To this end, we make 3 remarks:
\vspace{-5mm}
\begin{enumerate}
  \item First, the $\delta>0$ in the definition of SDIC is independent of $u$. 
  \item Second, in every neighborhood of $u$, we may not necessarily find all points $v$ in the neighborhood distinct from $u$ that would separate from the forward iterates of $u$. 
  \item  Finally, $N$ depends upon $u$ and $v$ chosen, and their iterates may not separate for ever (i.e., for all $n>N$) and we allow their iterates to get arbitrarily close in the future. 
\end{enumerate}

@@

\begin{Example}
  The logistic map(LM), a recursion relation of the form $x_{n+1}=rx_n(1-x_n)$ where $x_n\in[0,1]$ and $r<4$ is of particular interest to us as it is classically used to illustrate the influence of chaos on a system \textbf{(Cite)}. Below are plotted the first 75 iterates of the LM for $r=3.7$ and $x_0=0.2$ with an adjustment of 0.001 in $x_0$ and $r$ respectively.

  \begin{figure}[ht]
    \includegraphics[scale=0.45]{_log_sdic.eps}
        \centering
        \label{fig:log_sdic}
        \captionof{figure}{Figure generated for Logistic Map for $r=3.7$, $x_0=0.2$ where each parameter is varied separately by 0.001.}
      \end{figure}

  For an intuitive sense of the fact that the LM exhibits SDIC, we draw two green neighbourhoods of diameters $\delta_1=0.1, \delta_2=0.02$ around the points $u_1=0.7, u_2=0.4$ respectively. It is easy to see that every neighbourhood around these two points will contain some $v_1, v_2\in[0,1]$ and natural number $N_1, N_2 \in\mathbb{N}$ respectively s.t. the distance $d(T^Nu_1,T^Nv)>\delta_1=0.02$ and $d(T^Nu_2,T^Nv)>\delta_2=0.1$.

    \ednote{B: Is it desirable to have a more rigourous explanation here? }

\begin{Example}
  A second example pertains to a simple physical system called the double pendulum - a pendulum with another pendulum attached to its head given an initial position and angular velociy.

  Below are two figures denoting the trajectories of a double pendulum's second head after some elapsed time. Depicted are three pendulum heads with equal angular velocity, but differing ever so slightly in initial position . As can be seen below, the trajectories diverge very quickly to become completely different. 
  The double pendulum will be discussed in much greater detail in a later chapter (\ref{ch5}), but we mention the example here to provide a numerical example of a practical system exhibiting SDIC.

@@

Topologically transitivity implies that the iterates of an open set of initial conditions gets mixed up with other open sets. On a compact metric space, it can be shown that topological transitivity also implies the existence of a point whose forward iterates are dense \cite{de2013elements}; or in other words, the orbit going through this point will be dense in the compact metric space. 
In fact, it is topologically more likely that the choice of an arbitrary point will be one  whose iterates are almost dense.

@@

To be more precise (see \cite{de2013elements}), for a discrete-time dynamical system on a compact space, the set of points with dense iterates are residual, and they are typical or likely to be observed in practice. In this thesis, we assume that when data is observed from a topologically transitive system, we assume that the data arises from a dense orbit.

@@

\begin{Example} 
  The standard tent map $Tu=1-|2u-1|$ defined on $[0,1]$ is a well-known example of a dynamical system satisfying these three properties.

@@

To show topological similarity or sameness between two metric or topological spaces, one needs to establish a homeomorphism between the two spaces. 
However, in the study of dynamical systems defined on two spaces, establishing a homeomorphism does not indicate that the systems are dynamically related in any way.  For instance, the maps $Tu=u^2$ and $Tu=1-|2u-1|$ defined on $[0,1]$ have totally different behaviour. 
To find systems that are dynamically similar, one rather needs to establish a kind of dynamical equivalence which we illustrate in a simple commutativity diagram below.

@@

%Composite functions
To understand the diagram we note that if we travel ``right and then down," the diagram instructs us to use $T$ (top arrow right) first, followed by $\phi$. (right arrow downwards). Consequently the pathway amounts to finding $\phi(T(u))$. If we proceed "down and then right", the diagram instructs us to apply $\phi$ (left arrow downwards) first, and then apply $S$ (bottom arrow right). This then amounts to finding  $S(\phi(u))$. When the relationships in this diagram hold, we say $\phi(T(u))= S(\phi(u))$, and we formally denote it as $\phi \circ T=S\circ \phi$.

@@

\ednote{M: Saying $\phi$ is then a semi-conjugacy between $T$ and $S$ could be ambiguous as it does not specify which system is on the top in the commutativity. So please edit accordingly. B: Does defining explicitly the domain and codomain of $\phi$ make it clear enough? }
When $S$ is conjugate to $T$, the dynamics of the two systems are in some way ''dynamically equivalent''. Specifically, they are in one-to-one correspondence with one another. However when $S$ is semi-conjugate to $T$ with $\phi:U\to{V}$ a many-to-one mapping, the dynamics on $V$ provide merely a coarse-grained description of the dynamics on $U$ \cite{de2013elements}. When $S$ is semi-conjugate to $T$, it is also common to call $S$ a \emph{factor of $T$}, or conversely that $T$ is an \emph{extension of $S$}. In essence, an extension (e.g., \cite{de2013elements}) is a larger system capturing all of the important dynamics of its factor.

@@

In establishing that two systems are conjugate (semi-conjugate) to one another, we have also shown that one may choose to work with one system as opposed to the second and still be guaranteed to obtain information on the latter. This is especially useful while forecasting the future evolution of dynamical systems.

@@

In this chapter we acquaint ourselves with the question of forecasting dynamical systems with unknown underlying structures, state and discuss the shortcomings of the Takens delay embedding theorem and discuss various issues faced while forecasting.

@@

A practical example of this would be the subsequent scenario: given the time-sequential coordinates of an object moving in space, predict the future positions of that object. However we are very rarely, if at all, presented with a problem where the entire state information is available to us in the form $u_n$ at some time-step $n\in\mathbb{N}$. Consider then a more involved learning problem:

@@

First we establish the method of Takens delay embedding. 
In this method, one considers a discrete-time dynamical system defined on a 'nice' space (a smooth manifold)\ednote{B: Should this be further explained?} that can be obtained as time-$K$ map of a continuous time-dynamical system, concepts which will be formalised below.
We make observations from such a system, i.e., we observe the evolution of $\theta(w_0)$ (where $w_0$ represents the initial state) i.e., we observe  a finite set of values of $u_n :=  \theta(w_n) = \theta(Tw_{n-1})$. The sequence $\{u_n\}$ represents a scalar time-series and intuitively $\theta$ may be thought to represent a probe inserted into a larger system which is itself only measuring/extracting a small part of the greater system state at time $t=n$. 
Consider for example a thermometer erected to measure the ambient temperature in a local village. This measurement function, the thermometer, is capturing only a single aspect, the temperature, of a much grander dynamical system entailing the present weather of the surrounding area. Even more than that though- it is measuring a miniscule part of the global weather system.

@@

However, the true state $u$ of a system is seldom, if ever, fully known. In almost all cases we can at most insert probes into a system to obtain partial information by means of the measurements taken. Moreover, the process of taking a measurement itself introduces 2 additional aspects which complicate the problem: \ednote{M: Maybe replace difficulty by saying that we need to consider two aspects. B: Perhaps this edit makes more sense?}
\vspace{-8mm}
\begin{enumerate}[noitemsep, label=\roman*.]
  \item A series of measurements over a specific time-interval is inherently a discretisation procedure of the underlying continuous-time dynamical system. (Hence why we restricted our attention to discrete-time systems in the preceding section)
  \item A probe will never be fully accurate, and so the act of measurement introduces a certain measure of numerical noise/inaccuracy.
\end{enumerate}

From here we construct a multidimensional observable using the method of stacking previous observations, i.e., we create delay-coordinate map defined by
$\Phi_{k,\theta}(w) := (\theta(T^{-k}w)\ldots,\theta(T^{-1}w),\theta(w))$.  
The concept of an observable is understood in the sense of an observable as originally introduced by \cite{takens1981detecting, genericObservableAeyels} to refer to a probe or measurement function inserted into the system. 
The essence of Takens' theorem relates to the fact that when $k$ is sufficiently large, we can define a dynamical system on the space $\mathbb{R}^{k+1}$ whose states are $\Phi_{k,\theta}(w)$, $\Phi_{k,\theta}(Tw)$, $\Phi_{k,\theta}(T^2w)$, etc. and this dynamical system is topologically conjugate to the unknown underlying system $(W,T)$. We recall the Takens delay embedding theorem next.

@@

It is a fact that Takens' Embedding Theorem is both a powerful result and that it provides compelling reason to believe that one could conceivably reconstruct accurately a system conjugate to the underlying system. Nonetheless, it does present some serious practical limitations:
\vspace{-5mm}
\begin{enumerate}
\item Even supposing that we could indeed find $F_\theta$ , our \emph{approximation} of  $F_\theta$ is a map from a larger set $\mathbb{R}^{2d+1}$ containing the embedded attractor. There are, however, no theoretical guarantees that $F_\theta$ will retain  $V=\Phi_{2d,\theta}(W)$  as an attractor although $W$ could itself be an attractor.
\item Takens theorem is stated only for noiseless observations. Due to noise $\epsilon_n$ the delay vector $\Phi_{2d,\theta}(w_n) + \epsilon_n$  may lie outside $V$.Furthermore, due to the chaotic nature of the underlying system (i.e. the fact that it has SIDC), the evolution of $\Phi_{2d,\theta}(w_n) + \epsilon_n$ under the map $F_\theta$ could move out of $V$ completely. This problem can be overcome  by using a driven dynamical system with some properties, and we discuss this in Chapter\ref{ch3}. 
\end{enumerate}

The Takens embedding theorem does not guarantee global dissipativity.
\ednote{B:Links with~\ref{subs_LearnGamma} in discussion of dissipativity of $\Gamma$.  Sir, I'm not sure how to write this here - I think it a beneficial point to make so as to set the stage for the later discussion in ~\ref{subs_LearnGamma}, but I am unsure there as well.}

Though not a uniquely related to the Takens embedding theorem, we do opt to make mention of this requirement for our work. 
The methods described here are applicable for data originating from a surjective map. 
This follows from the fact that when $T$ maps the space $U$ into a proper subset $B$ of $U$, we me conclude that $T$ is not surjective as there exists some point in $U-A$ which does not have a preimage.
An example of this would be the case when a system displays energy-loss. Consider, for instance, consider the oscillations dying down for a damped pendulum. (A more in-depth account will be given in \ref{ch5})

@@

In this chapter, we discuss results pertaining to the mapping of temporal data obtained from a discrete-dynamical system onto a different space through the notion of a driven dynamical system. 
We also consider the conditions a driven system should possess to avoid adding distortion to its state-space representation. We then describe the \emph{single-delay dynamics} (SDD) of the system and and finally recall conditions on the driven system so that the SDD are conjugate (or at least semi-conjugate) to the underlying system. 
The SDD can then be used to forecast and reconstruct the underlying system.

@@

To find a function $\phi$, the need now arises to consider a so-called driven dynamical system, one that can generate a  nonautonomous dynamical system, where two spaces are considered: input is taken from an input space $U$ and the state of a space $X$ is updated. (Previously, only the state was considered.)
We do this to mimic the true conditions. By accounting for an extrogenous input sequence in $U$ which can also influence the  system-state at a specific time-step, it produces more general models than an autonomous system.

@@

The dynamics on X are generated by the update equation $x_{n+1}=g(u_n, x_n)$ where $n\in\mathbb{Z}$, input $u_n$ from $U$ and state $x_n$ belonging to $X$, where $X$ is compact.
Abbreviated, we shall refer only to the \emph{driven system $g$}, with all other entities being understood implicitly.

>>>>

Experimenting with biological, physical, and artificial systems to generate a more informative dynamical model is a well-established practice in modern science. Traditional methods for modelling physical systems are based on laws of physics that are based either on empirical relationships or intuition. For systems that evolve with time, physical laws yield mathematical equations that govern how the quantities evolve with time. However, our world is much more complex than that which can be distilled into elegant equations. We do not fully understand many of the more complex systems, nor do they provide us with a good physical intuition of the underlying principles governing system dynamics. To complicate this, the underlying systems we observe often display sensitive dependence on initial conditions despite having highly similar initial conditions. This is because differing orbits diverge quickly and to such an extent that it becomes seemingly impossible to retrace their steps back to the original conditions. Even the presence of usually' negligible' computational noise or measurement error renders long-term pointwise prediction infeasible. There are many difficulties that one encounters while modelling such systems:
\vspace{-8mm}
\begin{enumerate}[noitemsep, label=\roman*.]
  \item One may not have access to the complete states of the systems
  \item The system may be described by functions that behave wildly because their graphs have a wild oscillatory behaviour, i.e., a large functional complexity.
\end{enumerate}

@@

Examples in the literature attempting to forecast data from such systems have tried many different approaches with differing degrees of success. 
When we have complete access to states of the system, an ordinary differential equation can be obtained from data (\cite{brunton2016discovering, champion2019data} and \cite{small2002modeling,xu2006modeling}) wherein one could approximate the vector field by a library of functions to obtain interpretable models.

@@

This chapter briefly describes a discrete-time dynamical system and what it means to exhibit chaos. We refer to \cite{devaney2018introduction, de2013elements} for more details.

@@

Formally, a function $T : U \to U$, where U is some set is a discrete-time dynamical system and its iterates ${u, Tu, T^{2}{u},\ldots}$, where Tn denotes the n-fold composition of T with itself, describe the evolution of an initial condition $u \in U$ (Note that we frequently drop the brackets and represent $T(u)$ by $Tu$ to simplify notations).

@@

Of course, discrete-time dynamical systems need not always arise through a differential equation. Once again, we may consider the field of ecology, where discrete dynamical systems are often directly derived or assumed. Citation 
There is a school of thought that advocates discrete-time dynamical systems to be more natural for modelling real-world observations than differential equations. We refer the interested reader to \cite{saber2010introduction}.

@@

Two examples of invariant sets include a fixed point where $Tu=u$ for  $u\in U$, and a periodic orbit, i.e., a set of iterates $\{u,Tu, T^2u,\ldots,T^pu\}$, where $T^{p+1}u=u$ for some $p\in\mathbb{Z}$.  The entire space $U$ could also be invariant.
Consider for example the space $U=[0,1]$, where $Tu=4u(1-u)$ and then $U$ is invariant (as every $u\in{U}$ can be written as $u = 4x(1-x)$ for some $x\in{U}$).

@@

However, when the space U is compact, all orbits approach an invariant set. It follows because one can show the set of limit points of the orbit to be invariant~\cite{de2013elements}. So, when $U$ is compact,  the $\omega$-limit set $\omega(u;T)$ of a point $u$ defined to the collection of limit points of the sequence $\{x,Tu,T^2u,\ldots\}$ is nonempty, and $\omega(u;T)$ is invariant.

@@

Invariant sets have various properties. Invariant sets can be attracting or repelling depending on how orbits in their vicinity behave. 
Recall the examples Tu = u/2 and Tu = 2u defined on R (where {0} “attracts” orbits) or the example, Tu = u2 defined on [0,1] (where {1} “repels” orbits). 
We are interested in attractive invariant sets since they capture the long-term dynamics as time increases. 
In particular, we are interested in those invariant sets named attractors.

@@

Indeed in our previous examples, it can be verified that the system  $Tu=u/2$ defined on  $\mathbb{R}$ and $Tu=u^2$ defined on $[0,1]$  the singleton set $\{0\}$ is an attractor.  For the system,  $Tu=1-|2u-1|$ on $[0,1]$, one may easily verify that the only attractor is the entire space $[0,1]$. This follows from the fact that between any two points $u< v$ in $[0,1]$ and for any $a,b$ so that $u\le a < b \le v$, we can find an $n$ so that $T^n(a,b)=[0,1]$ (as would be explained later in this chapter).

A dynamical system can have several attractors and may also be contained in another attractor. For the example, $Tu = ^u2$ on $[0,1]$, both $\{0\}$ and [0,1] are attractors. It is known that an attractor for a dynamical system always exists in a compact space. \textbf{(Cite).}

The dynamics restricted to an invariant set can be complicated. For instance, an invariant set could be just a single point, or it could have an infinite set. If the invariant set is infinite, then complicated dynamics are possible. A particular well-studied phenomenon of such complexity gives rise to so-called chaotic behaviour, studied in detail over the past fifty years.

@@

In the 1960s, several mathematicians and mathematically interested scientists independently discovered chaos in the mathematical sense. The meteorologist Edward Lorenz may have been the first to explain this phenomenon in his 1963 paper [?]. The notions of invariance, attractivity, and chaos may also be described for continuous systems, and Lorenz's system comprised of a system of differential equations. The narrative of Lorenz's discovery of chaos and the history of other forerunners in this subject is fascinating. We highly recommend James Gleick's book Chaos: The Making of a New Science [?] for those interested. It clearly illustrates these experiences, explains why chaos was such a startling and crucial mathematical and scientific discovery and describes the underlying mathematical notions for non-specialists.

Several intricately different definitions of chaos are available in the literature (Cite), each indicating some aspect of complexity. In practice, it is not possible to verify which definition or notion the underlying dynamical system satisfies when only data is observed from it. However, we specifically recall the definition of chaos in the sense of Devaney [?, ?] to understand some nuances behind the complexity. Devaney's definition of chaos has three requirements, with the first being the notion of sensitive dependence on initial conditions.

@@

It is vital to acquire a sense of this definition as it can easily be misunderstood. It is a common misconception to interpret SDIC as two close points (u and v) that eventually become separated by a distance $\delta$ under iteration by T, but this is not true. To fully comprehend all the subtleties of the concept, one needs to discuss it more thoroughly by considering each sentence with care and attention; whereafter, one may examine how they fit together to convey the idea of sensitive dependence.

To this end, we make 3 remarks:
\vspace{-5mm}
\begin{enumerate}
  \item First, the $\delta>0$ in the definition of SDIC is independent of $u$. 
  \item Second, in every neighborhood of $u$, we may not necessarily find all points $v$ in the neighborhood distinct from $u$ that would separate from the forward iterates of $u$. 
  \item  Finally, $N$ depends upon $u$ and $v$ chosen, and their iterates may not separate forever (i.e., for all $n>N$) and we allow their iterates to get arbitrarily close in the future. 
\end{enumerate}

@@

\begin{Example} \rm
  The logistic map(LM), a recursion relation of the form $x_{n+1}=rx_n(1-x_n)$ where $x_n\in[0,1]$ and $r\leq{4}$ is of particular interest to us as it is classically used to illustrate the influence of chaos on a system \textbf{(Cite)}. 
  In the graph below~\ref{fig:log_sdic} we plot the first 50 iterates of the LM for $r=4$ and two slightly different initial values $x_0$.

  \begin{figure}[ht]
    \includegraphics[scale=0.74]{_logistic_sdic.eps}
        \centering
        \captionof{figure}{Figure generated for Logistic Map $x_{n+1}=rx_n(1-x_n)$ with $r=4$ to exhibit the presence of SDIC. Plotted are the values $x_n$ against time $n$ for the the first 50 step with initial values $x_0=0.2$ (blue) and $x_0=0.20001$ (red). Initially the two trajectories overlap, but they diverge completely at $n=12$ whereafter they follow distinct paths.}
        \label{fig:log_sdic}
      \end{figure}
  \ednote{B: Caption and figure adjusted}
  % For an intuitive sense of the fact that the LM exhibits SDIC, consider again the definition and substitute specific values. Let $\delta=0.05$ and for purposes of illustration consider the value $u=0.4$. The green band in~\ref{fig:log_sdic} shades the region $[u-\delta, u+\delta]$ and we can see that regardless the value $v$  and in every neighborhood of $u \in U$ there exists a $v\in{U}$ and an integer $N:=N{(u,v)}\in\mathbb{Z}$ such that $d(T^Nu,T^Nv)>\delta$
  % specific point, say $u=0.4$. Draw a green neighbourhood $[0.35,0.45]$ of diameters $\delta=0.02$ around the point. It is easy to see that every neighbourhood around $u=0.4$ will contain some $v_1, v_2\in[0,1]$ and natural number $N_1, N_2 \in\mathbb{N}$ respectively s.t. the distance $d(T^Nu_1,T^Nv)>\delta=0.05$ . 
\end{Example}

\begin{Example} \rm
  A second example pertains to a simple physical system called the double pendulum - a pendulum with another pendulum attached to its head given an initial position and angular velocity.

  Below are two figures denoting trajectories of a double pendulum's second head after some elapsed time. Depicted are three pendulum heads with equal angular velocity but differing slightly in initial position. As can be seen below, the trajectories diverge very quickly to become completely different. We discuss the double pendulum in greater detail in a later chapter (\ref{ch5}), but we mention the example here to provide a numerical example of a practical system exhibiting SDIC.

@@

Topological transitivity implies that iterates of an open set of initial conditions get mixed up with other open sets. On a compact metric space, one may show that topological transitivity also implies the existence of a point whose forward iterates are dense [?]; or in other words, the orbit going through this point will be dense in the compact metric space. 
In fact, it is topologically more likely that the choice of an arbitrary point will be one whose iterates are almost dense.

@@

To be more precise (see \cite{de2013elements}), for a discrete-time dynamical system on a compact space, the set of points with dense iterates are residual, and they are typical or likely to be observed in practice. In this thesis, when data is observed from a topologically transitive system, we assume it arises from a dense orbit.

@@

\begin{Example} \rm
  The standard tent map $Tu=1-|2u-1|$ defined on $[0,1]$ is a well-known example of a dynamical system satisfying these three properties.

@@

To show topological similarity (or sameness) between two metric or topological spaces, one needs to establish a homeomorphism between the two spaces. 
However, in the study of dynamical systems defined on two spaces, establishing a homeomorphism does not indicate that the systems are dynamically related in any way. 
For instance, the maps $Tu=u^2$ and $Tu=1-|2u-1|$ defined on $[0,1]$ have totally different behaviour. 
Therefore, to find dynamically similar systems, one must establish a dynamical equivalence that we illustrate in a simple commutativity diagram below~\ref{eqn_conjugacy}.

@@

%Composite functions
To understand the diagram, we note that if we travel "right and then down," the diagram instructs us to use T (top arrow right) first, followed by $\phi$. (right arrow downwards). Consequently, the pathway amounts to finding $\phi(T(u))$. If we proceed "down and then right", the diagram instructs us to apply $\phi$ (left arrow downwards) first, and then apply $S$ (bottom arrow right). This then amounts to finding  $S(\phi(u))$. When the relationships in this diagram hold, we say $\phi(T(u))= S(\phi(u))$, and we formally denote it as $\phi \circ T=S\circ \phi$.

@@

\ednote{B: Does defining explicitly the domain and codomain of $\phi$ make it clear enough?  M: Saying $\phi$ is then a semi-conjugacy between $T$ and $S$ could be ambiguous as it does not specify which system is on the top in the commutativity. So please edit accordingly.}
When $S$ is conjugate to $T$, the dynamics of the two systems are in some way ''dynamically equivalent''. Specifically, they are in one-to-one correspondence with one another. However when $S$ is semi-conjugate to $T$ with $\phi:U\to{V}$ a many-to-one mapping, the dynamics on $V$ provide merely a coarse-grained description of the dynamics on $U$~\cite{de2013elements}. When $S$ is semi-conjugate to $T$, it is also common to call $S$ a \emph{factor of $T$}, or conversely that $T$ is an \emph{extension of $S$}. In essence, an extension (e.g., \cite{de2013elements}) is a more extensive system capturing all of the essential dynamics of its factor.

@@

By establishing that two systems are conjugate (semi-conjugate) to one another, we have also shown that one may choose to work with one system instead of the second and still be guaranteed to obtain information on the latter. It is instrumental during the process of forecasting the future evolution of dynamical systems.

@@

In this chapter, we acquaint ourselves with the question of forecasting dynamical systems with unknown underlying structures, state and discuss the shortcomings of the Takens delay embedding theorem and discuss various issues faced while forecasting.

@@

A practical example of this would be the subsequent scenario: given the time-sequential coordinates of an object moving in space, predict the future positions of that object. However, we are very rarely, if at all, presented with a problem where the entire state information is available to us in the form un at some timestep $n\in\mathbb{N}$. Consider then a more complicated learning problem:

@@

First we establish the method of Takens delay embedding. 
In this method, one considers a discrete-time dynamical system defined on a 'nice' space (a smooth manifold)\ednote{B: Should this be further explained?} that can be obtained as time-$K$ map of a continuous time-dynamical system, concepts formalised below.
We make observations from such a system, i.e., we observe the evolution of $\theta(w_0)$ (where $w_0$ represents the initial state) i.e.\ednote{two i.e.'s}, we examine  a finite set of values of $u_n :=  \theta(w_n) = \theta(Tw_{n-1})$. The sequence $\{u_n\}$ represents a scalar time-series and intuitively $\theta$ represents a probe inserted into a bigger system which is itself only measuring/extracting a small part of the greater system state at time $t=n$. 
Consider, for example, a thermometer erected to measure the ambient temperature in a local village. This measurement function, the thermometer, is capturing only a single aspect, the temperature of a much grander dynamical system entailing the current weather of the surrounding area. Even more than that, though- it is measuring a miniscule part of the global weather system.

@@

However, the actual state $u$ of a system is seldom, if ever, fully known. In almost all cases, the most one can do is to insert probes into a system to obtain partial information through the measurements taken. Moreover, the process of taking a measurement itself introduces two additional aspects which complicate the problem: \ednote{M: Maybe replace difficulty by saying that we need to consider two aspects. B: Perhaps this edit makes more sense?}
\vspace{-8mm}
\begin{enumerate}[noitemsep, label=\roman*.]
  \item A series of measurements over a specific time interval is inherently a discretisation procedure of the underlying continuous-time dynamical system. (Hence why we restricted our attention to discrete-time systems in the preceding section)
  \item A probe will never be entirely accurate, and so the act of measurement introduces a certain measure of numerical noise/inaccuracy.
\end{enumerate}

From here we construct a multidimensional observable using the method of stacking previous observations, i.e., we create delay-coordinate map defined by
$\Phi_{k,\theta}(w) := (\theta(T^{-k}w)\ldots,\theta(T^{-1}w),\theta(w))$.  
The concept of an observable is understood in the sense of an observable as originally introduced by~\cite{takens1981detecting, genericObservableAeyels} to refer to a probe or measurement function inserted into the system.
Takens' theorem, in essence,  refers to the fact that for a sufficiently large $k$, we can define a dynamical system on the space $\mathbb{R}^{k+1}$ with states $\Phi_{k,\theta}(w)$, $\Phi_{k,\theta}(Tw)$, $\Phi_{k,\theta}(T^2w)$, etc. This dynamical system is topologically conjugate to the unknown underlying system $(W,T)$. We recall the Takens delay embedding theorem next.

@@

Takens' Embedding Theorem is a powerful result, and it provides compelling reason to believe that one could conceivably reconstruct a system conjugate accurately to the underlying system. Nonetheless, it does present some severe practical limitations:
\vspace{-5mm}
\begin{enumerate}
\item Even supposing that we could indeed find $F_\theta$ , our \emph{approximation} of  $F_\theta$ is a map from a larger set $\mathbb{R}^{2d+1}$ containing the embedded attractor. There are, however, no theoretical guarantees that $F_\theta$ will retain  $V=\Phi_{2d,\theta}(W)$  as an attractor although $W$ could itself be an attractor.
\item Takens theorem is stated only for noiseless observations. Due to noise $\epsilon_n$ the delay vector $\Phi_{2d,\theta}(w_n) + \epsilon_n$  may lie outside $V$.Furthermore, due to the chaotic nature of the underlying system (i.e. the fact that it has SIDC), the evolution of $\Phi_{2d,\theta}(w_n) + \epsilon_n$ under the map $F_\theta$ could move out of $V$ completely. This problem can be overcome  by using a driven dynamical system with some properties, and we discuss this in Chapter\ref{ch3}. 
\end{enumerate}

% The Takens embedding theorem does not guarantee global dissipativity.
% \ednote{B:Links with~\ref{subs_LearnGamma} in discussion of dissipativity of $\Gamma$.  Sir, I'm not sure how to write this here - I think it a beneficial point to make so as to set the stage for the later discussion in ~\ref{subs_LearnGamma}, but I am unsure there as well.}

Though not a uniquely related to the Takens embedding theorem, we do opt to make mention of another requirement for our work. 
The methods described here are applicable for data originating from a surjective map. 
This follows from the fact that when T maps the space U into a proper subset B of U, there exists some point in $U - A$ that does not have a preimage and we conclude that T is not surjective. 
An example of this would be the case when a system displays energy-loss. Consider, for instance, the dying oscillations for a damped pendulum. (A more in-depth account will be provided in 5)

@@

In this chapter, we discuss results on the mapping of temporal data obtained from a discrete-dynamical system onto a different space through the notion of a driven dynamical system. 
We also consider the conditions a driven system should possess to avoid adding distortion to its state-space representation. We then describe the single-delay Dynamics (SDD) of the system and finally recall conditions on the driven system so that the SDD are conjugate (or at least semi-conjugate) to the underlying system. 
The SDD can then be used to forecast and reconstruct the underlying system.

@@

To find a function $\phi$, the need  now arises to consider a so-called driven dynamical system, a special case of a nonautonomous dynamical system. Two spaces are considered: input is taken from an input space U and the state of a space X is updated. (Previously, only the state was considered.)
We do this to mimic the true conditions. By accounting for an extrogenous input sequence in $U$ which can also influence the  system-state at a specific time-step, it produces more general models than an autonomous system.

@@

The update equation $x_{n+1} = g(u_n,x_n)$ for $n \in\mathbb{Z}$, input $u_n$ from $U$ and state $x_n$ belonging to $X$ (a compact space) generate the dynamics on X. 
Abbreviated, we refer only to the \textit{driven system g} where all other entities are understood implicitly.
